```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

## Linear Regression

Before doing GWAS, we're going to learn about using linear models in R.

Loading a datset:

```{r, echo = FALSE}
data("mpg")
summary(mpg)
```

There is now an object in memory called `mpg`, which is a dataframe with 11 variables.

The mpg of cars in a city and the mpg on the highway are encoded in the columns `cty` and `hwy`, respectively. 

First, plotting these out: 

```{r}
ggplot(mpg, aes(x = cty, 
                y = hwy)) + 
  xlab("City Mileage (mpg)") + 
  ylab("Highway Mileage (mpg)") +
  geom_point() +
  theme_classic()
```

We can use the function `lm()` to implement a linear model. 

```{r}
lm(formula = hwy ~ cty, data = mpg)
```

This first argument, `formula` is what determines the variables we are regressing, with the tilde (~) sign separating dependent and independent variables. For example, the above formula asks to create a linear model where highway mileage is expressed as a function of city mileage. In other words, we're doing the good old algebra $$y = mx + b$$ except here it's $$highway = m * city + Intercept$$.

We can extract the intercept and coefficient as such: 

```{r}
regression <- lm(formula = hwy ~ cty, data = mpg)
print(regression$coefficients[1])
print(regression$coefficients[2])
```

Let's add this to our plot:

```{r}
ggplot(mpg, aes(x = cty, 
                y = hwy)) + 
  xlab("City Mileage (mpg)") + 
  ylab("Highway Mileage (mpg)") +
  geom_point() +
  theme_classic() +
  geom_abline(slope = regression$coefficients[2], 
              intercept = regression$coefficients[1])
```

Lastly, let's get some p values out from this. First, we get a summary of our model: 
```{r}
regression <- lm(formula = hwy ~ cty, data = mpg)
sumRegression <- summary(regression)
print(sumRegression)
```

From the `Coefficients`, we want to get the value of `Pr(>|t|)`. We can access `Coefficients` using the `$` operator: 

```{r}
print(sumRegression$coefficients)
```


And now we can index this to get our p value: 

```{r}
print(sumRegression$coefficients["cty", "Pr(>|t|)"])
```

This is a very low p value, reflecting the strongly non-zero slope of our regression line. 