```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

## Comparisons

In R, comparisons will always return Logical data, i.e. either `TRUE` or `False`. We can use the following syntax to compare values:

• `<` and `>`: greater than, less than \
• `<=` and `>=`: greater than or equal to, less than or equal to 

For example:

```{r}
print(5.3 < 17)
```

```{r}
print(4 >= (8/2))
```

• `==` checks if two values are equal. `!=` checks if two values are different.

```{r}
print(2 == (10 - 8))
```

Note that a single equals sign `=` can be used to assign values. However, a double equals sign `==` is used to compare values.

### AND and OR

We can evaluate multiple conditions using the logical AND and OR operators. 

#### AND

AND statements are represented using the operator `&`. Two AND statements evaluate as `TRUE` only if **both** are `TRUE`. For example:

```{r}
(5 > 2) & (7 + 2 == 9)
```

If **either** statement is `FALSE`, the entire expression is `FALSE`:

```{r}
(5 > 2) & (12 == 2)
```

#### OR

OR statements are represented using the operator `|`. Two OR statements evaluate as `TRUE` if **either** statement is `TRUE`. For example:

```{r}
(5 > 2) | (7 + 2 == 9)
```

If **either** statement is `TRUE`, the entire expression is `TRUE`:

```{r}
(5 > 2) | (12 == 2)
```