```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

## Plotting

The base R programming language comes with some pretty robust tools for plotting. However, most R users instead use the external library `ggplot2`, which simplifies plotting and allows for more elaborate data visualization. 

Your Posit cloud accounts have `ggplot2` pre-installed. However, at the start of each script where you plot, you must let R know that you will use an external package with the line:

```{r}
library(ggplot2)
```

For a simple dataset to plot, I’ve saved the temperature in Fahrenheit in Baltimore, MD and Wommels, Netherlands for the next seven hours:

```{r}
temperatures = data.frame(
  hours = seq(1,7),
  temperature = c(74, 73, 72, 71, 75, 79, 87, 59, 60, 62, 65, 69, 72, 67),
  city = c(rep("Baltimore", 7), rep("Wommels", 7))
)

head(temperatures)
```

```{r}
ggplot(temperatures)
```

```{r}
ggplot(temperatures, aes(x = hours, y = temperature))
```

```{r}
ggplot(temperatures, aes(x = hours, y = temperature)) + 
  geom_point() 
```

```{r}
ggplot(temperatures, aes(x = hours, y = temperature, color = city)) + 
  geom_point() 
```

```{r}
ggplot(temperatures, aes(x = hours, y = temperature, color = city)) + 
  geom_point() + 
  geom_line()
```

```{r}
ggplot(temperatures, aes(x = hours, y = temperature, color = city)) + 
  geom_point() + 
  geom_line() +
  xlab("Time (Hours)") + 
  ylab("Temperature (F)") +
  ggtitle("Projected Temperatures for the next Six Hours")
```

```{r}
ggplot(temperatures, aes(x = hours, y = temperature, color = city)) + 
  geom_point() + 
  geom_line() +
  xlab("Time (Hours)") + 
  ylab("Temperature (F)") +
  ggtitle("Projected Temperatures for the next Six Hours") + 
  theme_classic()
```

```{r}
ggplot(temperatures, aes(x = hours, y = temperature, color = city)) + 
  geom_point() + 
  geom_line() +
  xlab("Time (Hours)") + 
  ylab("Temperature (F)") +
  ggtitle("Projected Temperatures for the next Six Hours") + 
  theme_minimal()
```

```{r}
ggplot(temperatures, aes(x = hours, y = temperature, color = city)) + 
  geom_point() + 
  geom_line() +
  xlab("Time (Hours)") + 
  ylab("Temperature (F)") +
  ggtitle("Projected Temperatures for the next Six Hours") + 
  theme_bw()
```