```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

## Data Types

R allows us to work with many different types of data. The crucial ones for us are: 

**Numeric** data are all numbers. For example:

 • `78`\
 •  `3.2`\
 • `-9.8`
 
**Characters** are any text enclosed by single or double quotation marks. For example:

• `My Grandpa's deck has no pathetic cards`\
• `"85.3"`

Note that `85.3` - no quotes - is numeric data; `"85.3"` is character data.

**Logical** data has three possible values: `TRUE`, `FALSE`, and `NA` for missing data. These are case sensitive and lack quotes (`"TRUE"` would be character data). `TRUE` and `FALSE` can also be shortened to `T` and `F`. 

**Factors** are characters with associated **levels**, i.e. an order. For example, imagine that you have the set of characters `"Bach"`, `"Purcell"`, `"Handel"` and you always want them sorted in *that specific order*, rather than the default alphabetical order. We can use factors to manually create such an order for our data. If this is a bit abstract right now, it'll make more sense in practice once we work with real data!

## Variables

Variables allow us to store data in memory, associated with a name.

We create a variable using the assignment operator `<-` or `=`:

```
variable_name <- value
```

For our purposes, these are interchangeable. For example: 

• `composer <- "buxtehude"`\
• `year = 1637`

The variable name can be almost anything. Here are some general rules to consider when naming a variable:

• Variable names start with a letter and consist of letters, numbers, `.`, and `_`\
• Some words are considered "reserved" - i.e. they are already used by R to mean something and therefore cannot be a variable name. These include `TRUE`, `if`, `NULL`, etc. For a full list, check [here](https://www.datamentor.io/r-programming/reserved-words). \

### Overwriting Variables

In the following code example, we assign two different values to the same variable name:

```{r, eval = FALSE}
myData <- 3
myData <- "marigold"
```

What is the value of `myData`? It will be `"marigold"`. In the first line of code, we set `myData` equal to `3`, but then in the next line, we overwrite that value and set `myData` equal to `"margiold"`. The previous value of `myData` is erased and it is set to the new value. 

Note that in doing so, we change not just the value of `myData`, but also the *data type* - it goes from numeric to character data.

We can also do this to update the value of a variable:

```{r}
a_number <- 5 
a_number <- a_number + 10
print(a_number)
```

Here, we set the variable `a_number` equal to 5. In the next line, we add 10 to its value, storing the result under the same name. `a_number` is now equal to 15. 

