```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

## Vectors

Usually, we don't want to work with just a single data point - we will typically have multiple values that we want to store together. The most convenient way of doing this in R is using **vectors**. A vector stores multiple data points, preserving their order. 

We create a vector using the `c()` function. For example: 

```{r}
plants <- c("Feverfew", "Ivy", "Willow")
print(plants)
```

The data within a vector may be of any type, but all elements of a vector **must be of the same data type**. What happens if we try to create a vector with multiple data types? 

```{r}
beetles <- c("Weevil", "Firefly", 5)
print(beetles)
```

Here, we mix character and numeric data. Notice that when we print the vector, the numeric `5` is converted to the string `"5"`.

### Indexing

We will often want to take a larger vector and extract specific data points from it. To do this, we **index** our vector using the general syntax: 

`vectorName[itemPosition]`

The position of the first item in the list is 1, and so on (in other words, R is 1-indexed). Let's try indexing using our `plants` vector, made above

```{r}
print(plants)
```

To extract `"Ivy"`, we would do:

```{r}
plants[2]
```

We can also use a colon to extract multiple subsequent elements:

```{r}
plants[1:2]
```

We often want to extract elements near the end of a vector. `plants` is short and we can count to the end of it easily, but most of the data we will work with is a lot longer. One easy way to index items near the end of a vector is to use the `length()` function. We can index the final entry in plants as so:

```{r}
plants[length(plants)]
```

`length(plants)` is 3, so writing `plants[length(plants)]` is equivalent to writing `plants[3]`

Likewise, we can index the second element by doing some math: 

```{r}
plants[length(plants) - 1]
```

### Modifying Vecotrs


### Adding to Vectors








