```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

## Vectors

Usually, we don't want to work with just a single data point - we will typically have multiple values that we want to store together. The most convenient way of doing this in R is using **vectors**. A vector stores multiple data points, preserving their order. 

We create a vector using the `c()` (short for "concatenate") function. For example: 

```{r}
plants <- c("Feverfew", "Ivy", "Willow")
print(plants)
```

The data within a vector may be of any type, but all elements of a vector **must be of the same data type**. What happens if we try to create a vector with multiple data types? 

```{r}
beetles <- c("Weevil", "Firefly", 5)
print(beetles)
```

Here, we mix character and numeric data. This isn't allowed, so the numeric `5` is converted to the string `"5"`.

### Indexing

We will often want to take a larger vector and extract specific data points from it. To do this, we **index** our vector using the general syntax: 

`vectorName[itemPosition]`

The position of the first item in the list is 1 (in other words, R is 1-indexed). Let's try indexing using our `plants` vector, made above.

```{r}
print(plants)
```

To extract `"Ivy"`, we would do:

```{r}
plants[2]
```

We can also use a colon to extract multiple subsequent elements:

```{r}
plants[1:2]
```

We can also provide a vector to index multiple values:

```{r}
plants[c(1,3)]
```

We often want to extract elements near the end of a vector. `plants` is short and we can count to the end of it easily, but most of the data we will work with is a lot longer. One easy way to index items near the end of a vector is to use the `length()` function. We can index the final entry in plants as so:

```{r}
plants[length(plants)]
```

`length(plants)` is 3, so writing `plants[length(plants)]` is equivalent to writing `plants[3]`

Likewise, we can index the second element by doing some math: 

```{r}
plants[length(plants) - 1]
```

### Logical Indexing

We often want to subset our data not by the position of elements, but based on whether or not they meet a certain criterion. Below, I have generated a short list of numbers:

```{r}
myNumbers <- c(1, 54, 12.2, 70, 18, 24, 94)
```

Let's say we want to extract just the values that are greater than 15 from this list. We can use any of our comparative operators with a vector to compare *all* values within the vector:

```{r}
myNumbers > 15
```

This generates a logical vector. We can provide this vector as an index to `myNumbers` to pull out the elements that are `TRUE`.

```{r}
myNumbers[myNumbers > 15]
```

We can use the logical AND (`&`) and OR (`|`) operators to combine conditions. For example, extracting values greater than 15 and less than 30: 

```{r}
myNumbers[(myNumbers > 15) & (myNumbers < 30)]
```

### Modifying Vectors

Once we point to elements within a vector, we can modify them using the assignment operator. For example, making the second item in myNumbers equal to 200:

```{r}
myNumbers[2] <- 200
print(myNumbers)
```

We can also modify multiple elements at once. For example, making every value less than 50 equal to 0:

```{r}
myNumbers[myNumbers < 50] <- 0
print(myNumbers)
```

### Adding to Vectors

We can add to vectors using the concatenate function: 

```{r}
plants <- c(plants, "Philodendron")
print(plants)
```


