```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```


```{r, eval=FALSE}
library(ggplot2)
library(reshape2)
```

```{r, echo=FALSE}
library(ggplot2)
```

# Class 7: Infectious Diseases 

Before getting into any biology, let's learn about making our own functions. Throughout the course, we've made use of R functions such as `print()`, `c()`, and `ggplot()`; defining our own functions allows us to 

Defining our own functions allows us to perform actions over and over on multiple sets of inputs. 

Below, let's practice creating a function that takes one input, a list, and returns the number of things in it:

```{r}
# A vector containing the contents of my fridge 
refrigerator <- c("Apples", "Pears", "Bread", "Eggs", "Cheddar",
                  "Brie", "Lettuce", "Shallots", "Cabbage")

# The function definition 
countFood <- function(fridge){ # Fridge is a local variable - it does not exist outside the function 
  numItems <- length(fridge)
  print(paste("I have", numItems, "things in my fridge."))
}

# Two more fridges 
myFriendsPantry <- c("olives")
fridge <- c("veal", "steak", "halloumi")

# Run the fridge function on all three sets of inputs: 
countFood(refrigerator)
countFood(myFriendsPantry)
countFood(fridge)
```

The `countFood()` function prints out a value, but doesn't actually store anything to memory. Very often, we want to use the output from our function for downstream analysis. To save the oputput of our function, we need to include a `return()` statement.

For example, the function below generates a poly A tail sequence by pasting together a series of `"A"`'s, taking as input the length of the sequence to return: 

```{r}
# Define function
polyAtail <- function(length){
  polyA <- rep("A", times = length) # Create a vector of A's

  return(paste(polyA, collapse = "")) # Convert that vector into a string
}

# Run polyAtail() function on two inputs
polyA <- polyAtail(63)
polyA2 <- polyAtail(17)

# Print results
print(polyA)
print(polyA2)
```

### The SIR Model

In the function below, we implement the simplest form of the SIR model as :

```{r}
SIR <- function(S, I, R, # Starting number of individuals, as a fraction of total population size
                mu, B, gamma, # death/birth rate, contact lengths, recovery rate
                simTime, step){
  
  # Initialize total population size
  N <- 1
  
  # lists to store population sizes
  sList <- c(S)
  iList <- c(I)
  rList <- c(R)
  
  # loop through times
  for (time in seq(simTime/step)){
    # Calculate rates of change 
    
    dS <- (mu * (N - S) - B * I * S / N) * step
    dI <- (B * I * S / N - (mu + gamma) * I) * step
    dR <- (gamma * I - mu * R) * step
    
    # Update population sizes based on rate of change
    S <- S + dS
    I <- I + dI 
    R <- R + dR
    
    # Update lists
    sList <- c(sList, S)
    iList <- c(iList, I)
    rList <- c(rList, R)
  }
  # Store results to data frame
  df <- data.frame(
    time = seq(length(sList)) * step,
    S = sList, 
    I = iList,
    R = rList
  )
  
  return(df)
}
```

The function above returns a data frame containing time and population size for individuals of all categories.

We would also like to plot our data. Since we may want to do this over and over again for different inputs, it also makes sense to plot through a function. Notice that because we just want to create a plot here, we don't need to save anything to memory. Therefore, this function does not have a `return()` statement: 

```{r, eval=FALSE}
# Define plotting function 
plotSIR <- function(SIR_data){
  # Convert to tall format
  SIR_data_tall <- melt(SIR_data, id.vars = "time")
  # Fix the column names
  colnames(SIR_data_tall) <- c("Time", "Population", "Size")
  
  # Plot, coloring by population 
  ggplot(SIR_data_tall, aes(x = Time,
                            y = Size, 
                            color = Population)) + 
    geom_line() + 
    theme_bw()
}
```

We can also make a function to just plot the infected individuals. This is identical to the function above, except that 

```{r, eval=FALSE}
# Define plotting function 
plotInfected<- function(SIR_data){
  # Convert to tall format
  SIR_data_tall <- melt(SIR_data, id.vars = "time")
  # Fix the column names
  colnames(SIR_data_tall) <- c("Time", "Population", "Size")
  
  # Plot, coloring by population 
  ggplot(SIR_data_tall[SIR_data_tall$Population == "I", ], aes(x = Time,
                            y = Size, 
                            color = Population)) + 
    geom_line() + 
    theme_bw() 
}
```


Now that we've defined our function, we can run it on a set of data: 
```{r}
S <- 0.19
I <- 0.01
R <- 0.8
mu <- 1/(50 * 52)
B <- 2 # Contact lengths 
gamma <- 0.5 # Recovery rate 

simTime <- 52 * 100
step <- 0.1

SIR_data <- SIR(S, I, R, mu, B, gamma,simTime, step)
plotSIR(SIR_data)
```

```{r}
plotInfected(SIR_data)
```

```{r}
SIR_data_tall <- melt(SIR_data, id.vars = "time")
colnames(SIR_data_tall) <- c("Time", "Population", "Size")

ggplot(data=SIR_data, aes(x = S, y = I)) +
  geom_path() +theme_classic()
```

```{r}
SEIR <- function(S, E, I, R, # Starting number of individuals
                mu, B, gamma, sigma, alpha, p, # death/birth rate, contact lengths, recovery rate, disease progression rate, death rate from disease, vaccination proportion
                simTime, step){
  N <- 1
  
  # lists to store population sizes
  sList <- c(S)
  eList <- c(E)
  iList <- c(I)
  rList <- c(R)
  
  # loop through times
  for (time in seq(simTime/step)){
    # Calculate rates of change 
    
    dS <- (mu * (N * (1-p) - S) - B * I * S / N) * step
    dE <- (B * I * S / N - (mu + sigma) * E) * step
    dI <- (sigma * E - (mu + gamma + alpha) * I) * step
    dR <- (gamma * I - mu * R + mu * N * p) * step
    
    S <- S + dS
    E <- E + dE 
    I <- I + dI 
    R <- R + dR
    
    # Update lists
    sList <- c(sList, S)
    eList <- c(eList, E)
    iList <- c(iList, I)
    rList <- c(rList, R)
    
    N <- sum(S, E, I, R)
  }
  # Store results to data frame
  df <- data.frame(
    time = seq(length(sList)) * step,
    S = sList, 
    E = eList, 
    I = iList,
    R = rList
  )
  
  return(df)
}
```

```{r}
S <- 0.19
E <- 0.0
I <- 0.01
R <- 0.8
mu <- 1/(50 * 52)
B <- 2 # Contact lengths 
sigma <- 0.2
gamma <- 0.5 # Recovery rate 
alpha <- 0.1
p <- 0.2

simTime <- 52 * 50
step <- 0.1

SEIR_data <- SEIR(S, E, I, R, mu, B, gamma, sigma, alpha, p, simTime, step)
plotSIR(SEIR_data)
```

```{r}
ggplot(SEIR_data, aes(x = S, y = I)) +
  geom_path() + 
  geom_path(data = SEIR_data, aes(x = S, y = E), color = "plum") + 
  theme_classic()
```


















