---
title: "Infectious Diseases Drafts"
author: "Andrew Bortvin"
date: "2024-09-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(tidyverse)
library(reshape2)
library(statnet)
```

Implementation of the basic SIR model:

Closed epidemic (birth = death = 0)

```{r}
# Initialize simulation parameters
S <- 100 # Susceptible 
I <- 1 # Infected
R <- 0 # Recovered 
mu <- 0.00 # Birth rate, which also equals death rate 
B <- 3 # Infectiveness parameters - think of as duration of interactions
gamma <- 1 # Recovery rate 

N <- sum(S,I,R) # total population size 

simTime <- 30
step <- 0.01

# Lists to store values of S, I, R
sList <- c()
iList <- c()
rList <- c()

for (time in seq(simTime/step)){
  # Calculate rates of change
  
  # Change in number of susceptible 
  # First term: Births from all ages - deaths of S population
  # Second term: Fraction of S that become R
  dS <- (mu * (N - S) - B * I * S / N) * step
  
  # Change in number of infected
  # First term: Conversion of susceptible into infected (same as above)
  # Second term: Deaths and recoveries of infected individuals
  dI <- (B * I * S / N - (mu + gamma) * I) * step
  
  # Change in number of recovered
  # First term - number of recovered - note similarity but not complete identity with second term of I equation
  # Second term: Death of recovered individuals
  dR <- (gamma *I - mu * R) * step
  
  # Update population sizes
  S <- S + dS
  I <- I + dI
  R <- R + dR
  
  # Update population size 
  N <- sum(S,I,R)
  
  # Store to list
  sList <- c(sList, S)
  iList <- c(iList, I)
  rList <- c(rList, R)
}

# Store results in df
df <- data.frame(
  time = seq(length(sList)) * step,
  S = sList, 
  I = iList,
  R = rList
)

# Convert to tall format
df <- melt(df, id.vars = "time")
colnames(df) <- c("Time", "Population", "Size")

ggplot(df, aes(x = Time,
               y = Size,
               color = Population)) + 
  geom_line() +
  labs(y = "Population Size") +
  theme_bw() 
```

Defining as a function: 

```{r}
SIR <- function(S, I, R, mu, B, gamma, simTime, step){
  N <- sum(S,I,R) # total population size 
  
  # Lists to store values of S, I, R
  sList <- c()
  iList <- c()
  rList <- c()
  
  for (time in seq(simTime/step)){
    # Calculate rates of change
    
    # Change in number of susceptible 
    # First term: Births from all ages - deaths of S population
    # Second term: Fraction of S that become R
    dS <- (mu * (N - S) - B * I * S / N) * step
    
    # Change in number of infected
    # First term: Conversion of susceptible into infected (same as above)
    # Second term: Deaths and recoveries of infected individuals
    dI <- (B * I * S / N - (mu + gamma) * I) * step
    
    # Change in number of recovered
    # First term - number of recovered - note similarity but not complete identity with second term of I equation
    # Second term: Death of recovered individuals
    dR <- (gamma *I - mu * R) * step
    
    # Update population sizes
    S <- S + dS
    I <- I + dI
    R <- R + dR
    
    # Update population size 
    N <- sum(S,I,R)
    
    # Store to list
    sList <- c(sList, S)
    iList <- c(iList, I)
    rList <- c(rList, R)
  }
  
  # Store results in df
  df <- data.frame(
    time = seq(length(sList)) * step,
    S = sList, 
    I = iList,
    R = rList
  )
  
  return(df)
}
```


Running with a non-zero birth/death rate (open epidemic): 
```{r}
# Initialize simulation parameters
S <- 100 # Susceptible 
I <- 1 # Infected
R <- 0 # Recovered 
mu <- 0.05 # Birth rate, which also equals death rate 
B <- 3 # Infectiveness parameters - think of as duration of interactions
gamma <- 1 # Recovery rate 

simTime <- 40
step <- 0.001

SIR_data <- SIR(S, I, R, mu, B, gamma, simTime, step)

# Convert to tall format
SIR_data <- melt(SIR_data, id.vars = "time")
colnames(SIR_data) <- c("Time", "Population", "Size")

ggplot(SIR_data, aes(x = Time,
               y = Size,
               color = Population)) + 
  geom_line() +
  labs(y = "Population Size") +
  theme_bw() 
```

Running with a non-zero birth/death rate (open epidemic): 
```{r}
# Initialize simulation parameters
S <- 19 # Susceptible 
I <- 4 # Infected
R <- 0 # Recovered 
mu <- 0.05 # Birth rate, which also equals death rate 
B <- 2.76 # Infectiveness parameters - think of as duration of interactions
gamma <- 0.8 # Recovery rate 

simTime <- 50 
step <- 0.01

SIR_data <- SIR(S, I, R, mu, B, gamma, simTime, step)

# Convert to tall format
SIR_data <- melt(SIR_data, id.vars = "time")
colnames(SIR_data) <- c("Time", "Population", "Size")

ggplot(SIR_data, aes(x = Time,
               y = Size,
               color = Population)) + 
  geom_line() +
  labs(y = "Population Size") +
  theme_bw() 
```

## With Fractions: 

```{r}
SIR_fractions <- function(S, I, R, mu, B, gamma, simTime, step){
  N <- 1 # total population size 
  
  # Lists to store values of S, I, R
  sList <- c(S)
  iList <- c(I)
  rList <- c(R)
  
  for (time in seq(simTime/step)){
    # Calculate rates of change
    
    # Change in number of susceptible 
    # First term: Births from all ages - deaths of S population
    # Second term: Fraction of S that become R
    dS <- (mu * (N - S) - B * I * S / N) * step
    
    # Change in number of infected
    # First term: Conversion of susceptible into infected (same as above)
    # Second term: Deaths and recoveries of infected individuals
    dI <- (B * I * S / N - (mu + gamma) * I) * step
    
    # Change in number of recovered
    # First term - number of recovered - note similarity but not complete identity with second term of I equation
    # Second term: Death of recovered individuals
    dR <- (gamma * I - mu * R) * step
    
    # Update population sizes
    S <- S + dS
    I <- I + dI
    R <- R + dR
    
    # Update population size based on number dead
    N <- sum(S, I, R)
    
    # Store to list
    sList <- c(sList, S)
    iList <- c(iList, I)
    rList <- c(rList, R)
  }
  
  # Store results in df
  df <- data.frame(
    time = seq(length(sList)) * step,
    S = sList, 
    I = iList,
    R = rList
  )
  
  return(df)
}

```

```{r}
# Initialize simulation parameters
S <- 0.19 # Susceptible 
I <- 0.01 # Infected
R <- 0.8 # Recovered 
mu <-  1/(50 * 52) # Birth rate, which also equals death rate 
B <- 2 # Infectiveness parameters - think of as duration of interactions
gamma <- 0.5 # Recovery rate 

simTime <- 52 * 100
step <- 0.1

SIR_data <- SIR_fractions(S, I, R, mu, B, gamma, simTime, step)

# Convert to tall format
SIR_data_tall <- melt(SIR_data, id.vars = "time")
colnames(SIR_data_tall) <- c("Time", "Population", "Size")

ggplot(SIR_data_tall, aes(x = Time,
               y = Size,
               color = Population)) +
  geom_line() +
  labs(y = "Population Size") +
  theme_bw()



ggplot(SIR_data_tall[SIR_data_tall$Population == "I", ], aes(x = Time * step,
               y = Size,
               color = Population)) +
  geom_line() +
  labs(y = "Population Size") +
  theme_bw()
```

```{r}
ggplot(data = SIR_data, aes(x = S, y= I)) + 
  geom_path() + 
  theme_classic()
```

# SEIR

```{r}
SEIR_fractions <- function(S, E, I, R, mu, B, gamma, sigma, alpha, p, simTime, step){
  N <- 1 # total population size 
  
  # Lists to store values of S, I, R
  sList <- c(S)
  eList <- c(E)
  iList <- c(I)
  rList <- c(R)
  
  for (time in seq(simTime/step)){
    # Calculate rates of change
    
    # Change in number of susceptible 
    # First term: Births from all ages - deaths of S population
    # Second term: Fraction of S that become R
    # We still care about I here because I dictates the number of
    # individuals that get infected
    # We multiply N by (1-p) to reflect that only those not vaccinated at birth can be susceptible 
    dS <- (mu * (N * (1-p) - S) - B * I * S / N) * step
    
    # Change in number exposed
    dE <- (B * I * S / N - (mu + sigma) * E) * step
    
    # Change in number of infected
    # First term: Conversion of exposed into infected (same as above)
    # Second term: Deaths and recoveries of infected individuals
    dI <- (sigma * E - (mu + gamma + alpha) * I) * step
    
    # Change in number of recovered
    # First term - number of recovered - note similarity but not complete identity with second term of I equation
    # Second term: Death of recovered individuals
    # Third term: fraction of new births that are vaccinated - vaccinated births go directly into the R category 
    dR <- (gamma * I - mu * R + mu * N * p) * step
    
    # Update population sizes
    S <- S + dS
    E <- E + dE
    I <- I + dI
    R <- R + dR
    
    # Update population size based on number dead
    N <- sum(S, E, I, R)
    
    # Store to list
    sList <- c(sList, S)
    eList <- c(eList, E)
    iList <- c(iList, I)
    rList <- c(rList, R)
  }
  
  # Store results in df
  df <- data.frame(
    time = seq(length(sList)) * step,
    S = sList, 
    E = eList, 
    I = iList,
    R = rList
  )
  
  return(df)
}
```

```{r}
# Initialize simulation parameters
S <- 0.19 # Susceptible 
E <- 0.0  # Exposed
E <- 0.01  # Infected
R <- 0.8 # Recovered 
mu <-  1/(50 * 52) # Birth rate, which also equals death rate 
B <- 2 # Infectiveness parameters - think of as duration of interactions
sigma <- 0.2 # Length that exposed stay exposed prior to becoming infected
gamma <- 0.5 # Recovery rate 
p <- 0.2 # Vaccination rate at birth 
alpha <- 0.1 # Mortality rate of infected individuals

simTime <- 52 * 100
step <- 0.1

SEIR_data <- SEIR_fractions(S, E, I, R, mu, B, gamma, sigma, alpha, p, simTime, step)

# Convert to tall format
SEIR_data_tall <- melt(SEIR_data, id.vars = "time")
colnames(SEIR_data_tall) <- c("Time", "Population", "Size")

ggplot(SEIR_data_tall, aes(x = Time * step / 52,
               y = Size,
               color = Population)) +
  geom_line() +
  labs(y = "Population Size") +
  theme_bw()

```

```{r}
ggplot(SEIR_data_tall[SEIR_data_tall$Population == "I" | SEIR_data_tall$Population == "E", ], aes(x = Time * step,
               y = Size,
               color = Population)) +
  geom_line() +
  labs(y = "Proportion of Population Infected") +
  theme_bw()
```

```{r}
ggplot(data = SEIR_data, aes(x = S, y= R)) + 
  geom_path() + 
  theme_classic()
```

```{r}
ggplot(data = SEIR_data, aes(x = S, y= E)) + 
  geom_path() + 
  geom_path(data = SEIR_data, 
            aes(x = S, y = I),
            color = "plum") + 
  theme_classic()
```

```{r}
ggplot(data = SEIR_data, aes(x = S, y= E + I)) + 
  geom_path() + 
  theme_classic()
```

# SEIHFR

Model for Ebola transmission. Not a linear path from start to end - good to write this one out explicitly. Categories in the model:

- S: Susceptible
- E: Exposed
- I: Infected 
- H: Infected in hospital 
- F: Dead and not yet buried
- R: Removed (dead and buried, or immune)

```{r}
SEIHFR_fractions <- function(S, E, I, H, Funeral, R, # F spelled out due to protected terms 
                             Bs, Be, Bi, Bh, Bf, # All non-removed classes have beta parameters 
                             gammaH, gammaF, gammaR, # Progression rate from infection to hospitalization, death, or recovery
                             # 1/gamma is the time to each of those fates
                             theta, # Θ - proportion infected that are hospitalized
                             lambda, #Λ - proportion of the sick that will die
                             etaF, etaR,  # η - Progression from hospitalization to death or recovery
                             chi, mu, # time to burial
                             simTime, step){
  N <- 1 # total population size 
  
  # Lists to store values of S, I, R
  sList <- c(S)
  eList <- c(E)
  iList <- c(I)
  hList <- c(H)
  fList <- c(Funeral)
  rList <- c(R)
  
  for (time in seq(simTime/step)){
    # Calculate rates of change
    
    # Change in number of susceptible 
    # Interactions of susceptible individuals with Infected, Hospitalized, and dead individuals
    dS <- (mu * (N * (1-p) - S)-(Bi * I + Bh * H + Bf * Funeral) * S / N) * step
    
    # Change in number exposed 
    # First term: number of susceptible individuals that get exposed
    # Same as dS with opposite sign
    # Second term: number of exposed individuals that proceed to being infected
    dE <- ((Bi * I + Bh * H + Bf * Funeral) * S / N - sigma * E) * step
    
    # Change in number of infected
    # First term: number of exposed that become infected
    # Second term: progression from infection to hospitalization 
    # Third term: Recovery
    # Fourth term: Death
    dI <- (sigma * E - theta * gammaH * I - 
             (1 - theta) * (1 - lambda) * gammaR * I - 
             (1 - theta) * lambda * gammaF * I) * step
    
    # Change in the number hospitalized
    # First term: Progression from infection to hospitalization
    # Second term: Progression from hospitalization to death 
    # Third term: 
    dH <- (theta * gammaH * I - # infection to hospitalization
             lambda * etaF * H - # hospitalization to death
             (1 - lambda) * etaR * H) * step # hospitalization to recovery 
    
    # Change in number dead and unburied
    # First term: Progression from infection to death
    dFuneral <- ((1 - theta) * lambda * gammaF * I + # Infection to death. I think this is a typo in the textbook
                   lambda * etaF * H - # hospitalization to death
                   chi * Funeral
                 ) * step
    
    
    # Change in number of recovered
 
    dR <- ((1 - theta) * (1 - lambda) * gammaR * I + # infected recover
          (1 - lambda) * etaR * H # Recovery
          + chi * Funeral) * step # burial
    
    # Update population sizes
    S <- S + dS
    E <- E + dE
    I <- I + dI
    H <- H + dH
    Funeral <- Funeral + dFuneral
    R <- R + dR
    
    # Update population size based on number dead
    N <- sum(S, E, I, H, Funeral, R)
    
    # Store to list
    sList <- c(sList, S)
    eList <- c(eList, E)
    iList <- c(iList, I)
    hList <- c(hList, H)
    fList <- c(fList, Funeral)
    rList <- c(rList, R)
  }
  
  # Store results in df
  df <- data.frame(
    time = seq(length(sList)) * step,
    S = sList, 
    E = eList, 
    I = iList,
    H = hList,
    Funeral = fList,
    R = rList
  )
  
  return(df)
}
```

```{r}
# Initialize simulation parameters
S <- 0.99 # Susceptible 
E <- 0.0  # Exposed
I <- 0.01  # Infected
H <- 0
Funeral <- 0 
R <- 0.0 # Recovered 
Bi <- 0.588
Bh <- 0.794
Bf <- 7.653
gammaH <- 1/5
gammaF <- 1/9.6
gammaR <- 1/10
etaF <- 1/4.6
etaR <- 1/5
theta <- 0.80
lambda <- 0.81
chi <- 0.5
mu <-  1/(50 * 52)
sigma <- 1/7 # Length that exposed stay exposed prior to becoming infected


simTime <- 52 * 30
step <- 0.1

SEIHFR_data <- SEIHFR_fractions(S, E, I, H, Funeral, R, 
                 Bs, Be, Bi, Bh, Bf,
                 gammaH, gammaF, gammaR, 
                 theta,lambda, 
                 etaF, etaR,
                 chi, mu,
                 simTime, step)


# Convert to tall format
SEIHFR_data_tall <- melt(SEIHFR_data, id.vars = "time")
colnames(SEIHFR_data_tall) <- c("Time", "Population", "Size")

ggplot(SEIHFR_data_tall[SEIHFR_data_tall$Population != "R", ], aes(x = Time * step / 52,
               y = Size,
               color = Population)) +
  geom_line() +
  labs(y = "Population Size") +
  theme_bw()
```

```{r}
ggplot(SEIHFR_data_tall[SEIHFR_data_tall$Population == "I" | SEIHFR_data_tall$Population == "E" | SEIHFR_data_tall$Population == "H" , ], aes(x = Time * step,
               y = Size,
               color = Population)) +
  geom_line() +
  labs(y = "Proportion of Population Infected") +
  theme_bw()
```

```{r}
ggplot(data = SEIHFR_data, aes(x = S, y= R)) + 
  geom_path() + 
  theme_classic()
```

```{r}
ggplot(data = SEIHFR_data, aes(x = S, y= E + I + H)) + 
  geom_path() + 
  theme_classic()
```

# Networks

```{r}
ringlattice <- function(N, K){
  CM = toeplitz(c(0, rep(1, K), rep(0, N - 2 * K - 1), rep(1, K)))
  class(CM) = "cm"
  return(CM)
}
```

```{r}
plot.cm=function(CM){ 
  N=dim(CM)[1]
  theta=seq(0,2*pi,length=N+1) 
  x=cos(theta[1:N])
  y=sin(theta[1:N])
  symbols(x,y, fg=0, circles=rep(1, N),
          inches=0.1, bg=1, xlab="", ylab="") 
  segx1=as.vector(matrix(x, ncol=length(x),
                         nrow=length(x), byrow=TRUE)) 
  segx2=as.vector(matrix(x, ncol=length(x),
                         nrow=length(x), byrow=FALSE)) 
  segy1=as.vector(matrix(y, ncol=length(x),
                         nrow=length(x), byrow=TRUE)) 
  segy2=as.vector(matrix(y, ncol=length(x),
                         nrow=length(x), byrow=FALSE)) 
  segments(segx1,segy1, segx2, segy2, lty=as.vector(CM))
}
```

```{r}
cm = ringlattice(N = 20, K = 4)
plot(cm)
```

Barabasi-Albert Networks

```{r}
BarabasiAlbert <- function(N, K){
  CM=matrix(0, ncol=N, nrow=N) 
  CM[1,2]=1
  CM[2,1]=1
  for(i in 3:N){
    probs=apply(CM, 1, sum) 
    link=unique(sample(c(1:N)[-i],
                       size=min(c(K, i-1)), prob=probs[-i])) 
    CM[i, link]=CM[link, i]=1
    }
  class(CM)="cm" 
  return(CM)
}
```

```{r}
cm <- BarabasiAlbert(20, 4)
plot(cm)
```

```{r}
NetworkSIR=function(CM,tau,gamma){
  #generate SIR epidemic on a CM-network
  #CM = contact matrix
  #tau = probability of infection across an edge #gamma = probability of removal per time step 
  N=dim(CM)[1]
  I=matrix(rep(0,N),nrow=N,ncol=1) #First infecteds 
  S=matrix(rep(1,N),nrow=N,ncol=1) #First susceptibles 
  R=matrix(rep(0,N),nrow=N,ncol=1) #First removed 
  I1=sample(1:N, size=1)#Pick first random infected 
  I[I1,1]=1
  S[I1,1]=0
  t=1
  while(sum(I[,t-1])>0 | t==1){
      t=t+1
      infneigh=CM%*%I[,t-1]
      pinf=1-(1-tau)^infneigh 
      newI=rbinom(N, S[,t-1], pinf) 
      newR=rbinom(N, I[,t-1], gamma) 
      nextS=S[,t-1]-newI 
      nextI=I[,t-1]+newI-newR 
      nextR=R[,t-1]+newR
      I=cbind(I, nextI) 
      S=cbind(S, nextS) 
      R=cbind(R, nextR)
  }
  res=list(I=I,S=S,R=R) 
  class(res)="netSIR" 
  return(res)
}
```

```{r}
summary.netSIR=function(x){ 
  t=dim(x$S)[2] 
  S=apply(x$S,2,sum) 
  I=apply(x$I,2,sum)
  R=apply(x$R,2,sum) 
  res=data.frame(S=S,I=I,R=R) 
  return(res)
}

plot.netSIR=function(x){ 
  y=summary(x)
  plot(y$S, type="b", xlab="time", ylab="") 
  lines(y$I, type="b", col="red") 
  lines(y$R, type="b", col="blue") 
  legend("right", legend=c("S", "I", "R"),
         lty=c(1,1,1), pch=c(1,1,1),
         col=c("black", "red", "blue")) 
}
```

```{r}
cm <- BarabasiAlbert(N = 75, K = 3)
sim <- NetworkSIR(cm,.06,0.08)
plot.netSIR(sim)
```

```{r}
cmg <- cm
inf=ifelse(apply(sim$I, 1, sum)>0,2,1)
nwt=network(cmg, directed=FALSE)
plot(nwt, vertex.col=inf)
```

# Seasonality

```{r}
SIR_recurrence <- function(S, I, R, mu, B, gamma, omega, simTime, step){
  N <- 1 # total population size 
  
  # Lists to store values of S, I, R
  sList <- c(S)
  iList <- c(I)
  rList <- c(R)
  
  for (time in seq(simTime/step)){
    # Calculate rates of change
    
    # Change in number of susceptible 
    # First term: Births from all ages - deaths of S population
    # Second term: Fraction of S that become R
    dS <- (mu * (N - S) - B * I * S / N + omega * R) * step
    
    # Change in number of infected
    # First term: Conversion of susceptible into infected (same as above)
    # Second term: Deaths and recoveries of infected individuals
    dI <- (B * I * S / N - (mu + gamma) * I) * step
    
    # Change in number of recovered
    # First term - number of recovered - note similarity but not complete identity with second term of I equation
    # Second term: Death of recovered individuals
    dR <- (gamma * I - mu * R - omega * R) * step
    
    # Update population sizes
    S <- S + dS
    I <- I + dI
    R <- R + dR
    
    # Update population size based on number dead
    N <- sum(S, I, R)
    
    # Store to list
    sList <- c(sList, S)
    iList <- c(iList, I)
    rList <- c(rList, R)
  }
  
  # Store results in df
  df <- data.frame(
    time = seq(length(sList)) * step,
    S = sList, 
    I = iList,
    R = rList
  )
  
  return(df)
}
```

```{r}
# Initialize simulation parameters
S <- 0.99 # Susceptible 
I <- 0.01 # Infected
R <- 0 # Recovered 
mu <-  1/(50 * 52) # Birth rate, which also equals death rate 
B <- 4 # Infectiveness parameters - think of as duration of interactions
gamma <- 1 # Recovery rate 
omgea <- 0.03

simTime <- 52 * 3
step <- 0.1

SIR_data <- SIR_recurrence(S, I, R, mu, B, gamma, omgea, simTime, step)

# Convert to tall format
SIR_data_tall <- melt(SIR_data, id.vars = "time")
colnames(SIR_data_tall) <- c("Time", "Population", "Size")

ggplot(SIR_data_tall, aes(x = Time,
               y = Size,
               color = Population)) +
  geom_line() +
  labs(y = "Population Size") +
  theme_bw()



ggplot(SIR_data_tall[SIR_data_tall$Population == "I", ], aes(x = Time * step,
               y = Size,
               color = Population)) +
  geom_line() +
  labs(y = "Population Size") +
  theme_bw()

```

Ways to make this more exciting: 

* Periodicity in the infection rate? Seasonality? To what extent is this realistic? 
* See textobook p 29 for concrete explanation of R0 as a function of B and gamma
* Seasonal SEIR
* Age groups in SEIR

* Also - look into Lotka-Volterra parasitism

BI/N term in S(E)IR is the Force of Infection - per capita rate of exposure to infection 








































