```{r, include = FALSE}
ottrpal::set_knitr_image_path()
knitr::opts_knit$set(root.dir = '01-intro_to_R')
```

# Introduction to R 

R is a programming language developed primarily for use by statisticians, and it is a common tool in the field for data exploration, analysis, visualization, etc. We will be using it throughout the course to manipulate, view, and interpret data. 

In this module, we will learn the basics of R by manipulating some real-world data. In particular, we'll look at data from the Hudson Bay looking at the size of the hare and lynx populations. 

We'll talk a lot more about this dataset when we study predator-prey models in a few weeks, but for now, let's make some exploratory plots. 

Data adapted from here: http://people.whitman.edu/~hundledr/courses/M250F03/M250.html

Please note that input data is available on the course Posit cloud page - please feel free to work alongside this page! 

## Data import 

The bread and butter of R is the dataframe, a tabular data structure which contains data of multiple types. 

R has a few built-in functions that allow for easy and efficient import of data. We'll learn a lot more nuances of how they work as we go on, but for now, we can import a tidy, well-formatted bit of data as such: 

```{r, include = TRUE}
print("test")
```


## Data types

R has the ability to manipulate multiple different types of data. The primary data types we will use in this course are: 

**Numeric** are whole numbers. For example:

