```{r, include = FALSE}
ottrpal::set_knitr_image_path()
knitr::opts_knit$set(root.dir = '01-intro_to_R')
```

# Introduction to R 

R is a programming language developed primarily for use by statisticians, and it is a common tool in the field for data exploration, analysis, visualization, etc. We will be using it throughout the course to manipulate, view, and interpret data. 

In this module, we will learn the basics of R by manipulating some real-world data. In particular, we'll look at data from the Hudson Bay looking at the size of the hare and lynx populations. 

We'll talk a lot more about this dataset when we study predator-prey models in a few weeks, but for now, let's make some exploratory plots. 

Data adapted from here: http://people.whitman.edu/~hundledr/courses/M250F03/M250.html

Please note that input data is available on the course Posit cloud page - please feel free to work alongside this page! 

## Variables in R



## Data import 

The bread and butter of R is the dataframe, a tabular data structure which contains data of multiple types. 

R has a few built-in functions that allow for easy and efficient import of data. As we go on, we'll learn a lot of nuances about how these work, but for now we'll start with a tidy, well-behaved dataset that is easy to work with.

We would like for R to import our data and save it into memory. We also want to give it a name, so that we can refer to it later. 

In R, we use the assignment operator `<-` to associate a name with some data. The general syntax for this is: 

`variableName <- value`

Where **left - name right -> value**

**if you know python or c or something, "+" is equivalent**

**lets test it out**

```{r, include = TRUE}
populationData <- read.table("hare_lynx.csv", sep = ",", header = TRUE)
```

Let's break down what's happening here.

On the left of the `<-`, we have the text `populationData` 

```{r}
head(populationData)
```

## The Data Frame

R has the ability to manipulate multiple different types of data. The primary data types we will use in this course are: 

**Numeric** are whole numbers. For example:

