```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

# Lecture Notes

## Class 2: Intro to R

8/27/2024

Today I want to give a pretty broad overview of the basics of using R. There are three big topics that I want to hit upon today. The first is just the basics of the language - what are the standard ways we interface with R? What data can we work with? So on. 

Secondly, I want to introduce you to working with tabular data. Data tables are the heart and soul of R, and so today we'll learn how to make them, read them in from a file, and manipulate them.

And lastly, I want to teach you a bit of the basics of plotting in R. 

Every line of code that we write in an R script evaluates, unless we preface it with the hashtag `#`. In that case, everything in the line of code following the `#` is a comment and is ignored by the interpreter. 

We can run our script by clicking, or with the nice hotkey Shift+Command+Return (Windows: Shift+Control+Enter). We can run our individual lines with Command+Return (Windows: Control+Enter).

## Data Types, Variables, Math

### Characters and Variables

In R, we have access to a lot of different data types. Text data is called **Character data** and is denoted by single or double quotes:

```{r}
"ant"
```

We've just written down that string and not actually done anything with it. Typically, though, we want to save and store our data for future use. 

We can do this using **variables**. We assign a variable using the operators `<-` or `=`. For our purposes these are interchangeable for assignment; if you look at code online you'll likely see both. I use `<-` since it is a little more common. 

```{r}
insect <- "cicada"
```

Now we have a thing in memory with the name `insect`. And associated with that name is some data - in this case, the word "ant". Incidentally, you can see in the top right corner of your screen in the Environment tab there is now an `insect` variable with the associated word "ant".

And now, whenever we want to access that data, we can use its variable name. For example, the `print()` function allows us to, fittingly, print data to the screen. We can print the contents of `insect` as such:

```{r}
print(insect)
```

And so just one bit of nomenclature - here, `print()` is the **function**, and **insect** - the thing we give to the function - is the function's **argument**.

A little more about assignment: we can "overwrite" a variable by assigning new data to it. 

```{r}
insect <- "ant"
print(insect)
```

Now, `insect` is associated with the value `"ant"`. The old value is removed and the new one is added in its place. 

I've just told you that `insect` is character data, but if you ever are working with data and you don't know the type of data it is, you can investigate that using the `class()` function:

```{r}
class(insect)
```

#### paste

We have two very helpful functions for combining strings `paste` and `paste0`. 

```{r}
paste(insect, "hill")
```

We can give `paste()` a separator to combine with a common delimiter: 

```{r}
paste("eye", "of", "newt", sep = ",")
```

`paste0()` is a shorthand of writing `sep = ""`

```{r}
print(paste(insect, "hill", sep = ""))
print(paste0(insect, "hill"))
```

### Numbers

In many programming languages, you need to be very careful about different types of data - the difference between integers and numbers with decimal points is often extremely critical. R simplifies this a lot - under the hood, there are differences between these data types, but for our purposes we can broadly treat them as one type of data - **numeric data**.

Math is straightforward:

```{r}
aNumber <- 3.14
print(aNumber + 2)
```

Notice that this **does not** modify the original value of `aNumber`:

```{r}
print(aNumber)
```

To do that, we must use the assignment operator: 

```{r}
aNumber <- aNumber * 2
print(aNumber)
```

Other Math:

```{r}
print(5 + 2.2) # Notice the accommodation of integers and floating point data
print(7/3) # Division
print(9*4) # Multiplication
print(2**3) # Exponents
```

This last point about math is fairly intuitive, so I don't need to dwell on it too long, but R obeys standard order of operations. Just like in real math, you can use parenthesis to control the order of operations. 

```{r}
print(1 + 3 * 2)
print((1 + 3) * 2)
```

### Logical Data

R can represent whether conditions are true or false using the appropriately named, case-sensitive `TRUE` and `FALSE`. These can shortened to `T` and `F`. 

A really nifty part of these is that they can also be automatically converted to the numeric `1` for `TRUE` and `0` for `FALSE`. 

```{r}
T + TRUE + FALSE
```

Now that we've seen the three major data types, one last thing about the assignment operator. Assignment can change the type of a variable. For example:

```{r}
testData <- 5
print(class(testData))
testData <- "yes"
print(class(testData))
```

## Comparisons

Logical data most commonly appears when we compare values. We compare equality using the double equals sign:


```{r}
print("egg" == "egg")
print("Egg" == "egg")
```

We can invert this using the `!=` operator: 

```{r}
print(2 != 3)
```

We can likewise use `>` and `<` to check if things are greater or less than each other.

```{r}
print(5 > 2.2)
```


`>=` and `<=` are greater than/less or equal to: 

```{r}
print(3 <= 3)
```

We can also use the exclamation point `!` to negate logic in R: 

```{r}
print(!TRUE)
```

```{r}
print(!(3 > 2))
```

### Multiple Conditions

AND and OR operators are used to evaluate multiple conditions. Any number of conditions connected by AND and OR operators return only one `TRUE` or `FALSE` value.

#### AND

The logical AND is denoted by the operator `&`. Two conditions connected by `&` evaluate as `TRUE` only if **both** are `TRUE`:

```{r}
(5 + 2 ==7) & ('a' != 'b') # True and True
```

If a single condition is `FALSE`, the entire statement is `FALSE`:

```{r}
(3 + 2 < 4) & (5 > 2) # False and True
```

#### OR

The logical OR is denoted by the operator `|`. Two conditions connected by `|` evaluate as `TRUE` if **either** condition is `TRUE`:

```{r}
(7 >= 3) | (2 == 3) # True and False
```

## Collections

Most of the time, our data doesn't come in the form of individual discrete observations. Instead, we usually work with collections of data - for example, the size of a population across many years.  


The most basic kind of collection in R is the **vector**, which combines multiple data points of the same type. We create a vector using the function `c()` (stands for concatenate).

```{r}
fruits <- c("peach", "plum", "pear")
print(fruits)
```

If we mix data of multiple types, R automatically converts data, to the best of its ability: 

```{r}
fruits <- c("peach", 5.2, "pear")
print(fruits)
```

Notice that the numeric 5.2 is converted to the character `"5.2"` (see the quotation marks).

### Adding to vectors 

We can use the `c()` function to add to an existing vector:

```{r}
fruits <- c(fruits, "apple")
print(fruits)
```

### Indexing 

We can extract from a vector using the brackets `[]`. Inside the brackets, put the position of the item you want to extract: 

```{r}
print(fruits)
print(fruits[2]) # Get the second item
print(fruits[length(fruits)]) # Get the last item
```

We can provide a range or a vector of positions to extract multiple items from our vector: 

```{r}
print(fruits[2:4])
print(fruits[c(1,3,2)])
```

We can combine indexing and the assignment operator to modify a vector. For example, changing `"5.2"` back to `"plum"`:

```{r}
fruits[2] <- "plum"
print(fruits)
```

### Comparisons with vectors

We can use our comparison operators with vectors to compare the entire vector to a value. 

```{r}
myNumbers <- c(4, 10, 8, 2)
myNumbers > 5
```

We return a vector of logical values! 

### Two other major types of collections: 

#### Rep

We often want to repeat elements a set number of times. Writing this out by hand is annoying, so there's a helpful function to do it for you: `rep()`.

```{r}
rep(5, times = 4)
```

This repeats `5` four times. 

We can also repeat vectors: 

```{r}
rep(c('stop', 'go'), times = 4)
```

What if we want to repeat each element in the vector a set number of times before moving on to the next? e.g. we want to turn `c("a", "b")` into `c("a","a","a", "b","b","b")`?

```{r}
rep(c('a', 'b'), each = 3)
```

So, notice that this the first function we've encountered that takes multiple arguments. Multiple arguments are extremely common. For most functions, the first argument will be the input data or a formula or something. It doesn't need a name - R can figure that out on its own. 

After that, though, most arguments are **optional**. So if you give R the code `rep(c('stop', 'go'), 4)`, it won't automatically know what that 4 means - it could be either `times` or `each` in that position. So it's a good practice to explicitly tell R what your optional arguments are by writing their name. 

Incidentally, how do we know what these optional arguments can be? This is something we need to know on a function-by function basis, and the ones for the common functions will be second nature to you sooner than you think. But also, we can look up our functions documentation using `?rep`. And if our function is not from the default R language, we can do a broader search using `??rep?`

#### seq

We often want to work with a series of numbers in order (1,2,3,4,5,...,n). The `seq()` function allows us to do this. 

Let's look at it in a bit more detail:

```{r}
seq(5)
```

We can give `seq()` multiple numbers to create a start and stop condition:

```{r}
seq(1, 10)
seq(7,3)
```

Two important optional arguments for `seq()`

```{r}
seq(5,20, by = 5)
```

```{r}
seq(4,9, length = 12)
```

## Iteration

An extremely common thing for us is that we will want to perform actions over and over again. Say you have 30 individuals in a population and every generation it gets 10% bigger. We can write that out over and over, but it gets tiring and impractical soon. 

```{r}
for (i in seq(1, 5)){
  print(i)
}
```

So what we do here is take this sequence of 5 items. For each item, we take this temporary variable `i`, and we perform an action on it. Here, we just print it out. 

```{r}
for (i in seq(1, 5)){
  i <- 2**i
  print(i > 6)
}
```

We don't need to operate on `i` directly. For example, to implement that population example: 

```{r}
N <- 30
for (i in seq(1, 8)){
  N <- N * 1.1
  print(N)
}
```

So, here we take this variable N and we make it 10% bigger 8 times using our loop. 

The last thing I want to touch upon here is that we put a for loop inside a for loop:

```{r}

for (i in c("b", "c")){
  for (j in c("at", "ar")){
    print(paste0(i, j))
  }
}
```

## Data Frames 

So vectors are great, but usually we don't keep track of just *one* thing - we normally have multiple variables that we track simultaneously - for example, time and population size. A data frame is a tabular data structure that lets us organize a lot of different data in one place. 

Each column in a data frame is a vector - so within each column, the data needs to be of the same type. However, each column can be its own data type. So to make this concrete, let's build a simple data frame with some made up data:

```{r}
myDf <- data.frame(
  numbers = c(1, 4, 6, 2, 3.3),
  logic = c(T, T, F, F, T),
  words = c("red", "green", "blue", "orange", "turquoise")
)

myDf
```

So now that we've made a simple dataset, let's load in some more elaborate real data. 

```{r}
penguins <- read.csv("penguins.csv")
```

We have a few ways of quickly looking at our data: 

```{r}
head(penguins)
tail(penguins)
```

and `View(penguins)`, which is analogous to clicking on it in the Environment tab. 

So we can only see the first seven columns of our data here - if we ever want to know what all the columns are, we can use the `names()` function:

```{r}
names(penguins)
```

We can also pretty easily get the dimensions of our data:

```{r}
nrow(penguins)
ncol(penguins)
dim(penguins)
```

### Subsetting

We can extract columns from our data frame using the dollar sign. So for example if I want to see the `island` column, I can do:

```{r, eval = FALSE}
penguins$island
```

Which we can now index the usual way: 

```{r}
penguins$island[5:10]
```

We can also index the entire df the conventional way, but now we need to provide a row and column index:

```{r}
penguins[5, 10]
```

### Logical Indexing

Let's say that we want the penguins with big bills. 

We can use logical comparisons with the `bill_length_mm` vector to find which ones are 40 mm or longer.

```{r, eval=FALSE}
penguins$bill_length_mm >= 40
```

This creates a huge vector of `TRUE`s and `FALSE`s. We can give this as a row index to pull out the `TRUE` rows:

```{r, eval=FALSE}
penguins[penguins$bill_length_mm >= 40, ] 
```

Notice that the column index is blank - this means that we want to pull out all columns. 

Notice that we retain rows where the bill length is `NA` are retained. The easiest way to address this is to remove rows with `NA`s. Checking for `NA` values in R is a bit finicky. The conventional way is to use the function `is.na()`. This returns `TRUE` for all `NA`s. We want the opposite of that, so we add in `!` to negate it:

```{r, eval=FALSE}
penguins[penguins$bill_length_mm >= 40 & !is.na(penguins$bill_length_mm), ] 
```

We can also use this type of logical comparisons to answer the question of *how many* penguins have long bills. `penguins$bill_length_mm >= 40` returns a logical vector of `T` and `F`. Recall that these are equivalent to `1` and `0`. So to find the number of `TRUE`s, we can just sum the vector: 

```{r}
sum(penguins$bill_length_mm >= 40, na.rm = TRUE)
```

Again, we need to remove the `NA`s for this to work as intended, so the sum function has an optional `na.rm` argument. 

## Plotting 

For plotting, we use the external library `ggplot2`. This library is installed but because it is not a part of the base R programming language, we need to let R know that we will be using it. This is done with the following line of code: 

```{r}
library(ggplot2)
```

To create a plot, we need to provide a data set that we will plot, along with defining some aesthetics (axes, colors, groups of data, etc.). Almost always, we will need to define the x and y axes. Let's start by plotting the bill length and body mass of our penguins.

```{r}
ggplot(penguins, aes(x = body_mass_g, y = bill_length_mm))
```

This already creates a nice plot. Notice that the axes are already scaled around your data. 

Now we can add points. To add a geometric, we use the addition operator `+` to add a type of geometric object to our plots:

```{r}
ggplot(penguins, aes(x = body_mass_g, y = bill_length_mm)) +
  geom_point()
```

Now, we can also color our data by a different variable. This involves adding a `color` term to `aes` For example, recoloring by sex:

```{r}
ggplot(penguins, aes(x = body_mass_g, 
                     y = bill_length_mm,
                     color = sex)) +
  geom_point()
```

You may notice that some of our penguins are missing sex information. This is coded by an empty string `""` rather than an `NA`. We can filter these out as such: 

```{r}
penguinsFiltered <- penguins[penguins$sex != "", ] 
```

We retain just the rows where the `sex` column is not equal to `"`. Plotting this new dataset:

```{r}
ggplot(penguinsFiltered, aes(x = body_mass_g, 
                     y = bill_length_mm,
                     color = sex)) +
  geom_point()
```

Another way to separate variables is to create multiple subplots. This is mostly easy done by adding `facet_wrap()` to our `ggplot()` statement. To `facet_wrap()`, we give the name of the column that we want to split by. For example, splitting the above plot by island: `

```{r}
ggplot(penguinsFiltered, aes(x = body_mass_g, 
                     y = bill_length_mm,
                     color = sex)) +
  geom_point() + 
  facet_wrap("island")
```

Notice that for `facet_wrap()`, we need to give the column name in quotation marks.

## Wide to Long 

One thing you may have noticed is that when we plot our data, we want our data to be tall rather than wide. To make that a little more concrete, here's a data set of population sizes for two species:

```{r}
popSizes <- data.frame(
  generation = seq(1,5), 
  species1 = c(10, 12, 15, 13, 14),
  species2 = c(20, 15, 14, 13, 11)
)

head(popSizes)
```

So if we go to plot this, we hit an issue:

```{r, eval=FALSE}
ggplot(popSizes, aes(x = generation, y = #????))
```

There are a few workarounds for this, but one of the best ones is to reshape your data and make it taller. 

The `reshape2` library has a great function for this is: 

```{r, eval=FALSE}
library(reshape2)
```

```{r}
popSizesTall <- reshape2::melt(popSizes, id.vars = "generation")
head(popSizesTall)
```

We should now rename our columns to be a bit more descriptive. This can be done with the `colnames()` function:

```{r}
colnames(popSizesTall) <- c("generation", "species", "size")
```

Plotting is now straightforward: 

```{r, eval=FALSE}
ggplot(popSizesTall, aes(x = generation, y = size, color = species)) +
         geom_point() +
         geom_line() + 
         theme_classic()
```

